---
title: "Class09 Unsupervised learning"
author: "Pin-Chung (Tony) Cheng"
date: "10/30/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
wisc.df = read.csv("WisconsinCancer.csv")
head(wisc.df)

```

# `r ` >  cool way to put r code in sentence
We examine data from `r nrow(wisc.df)` patient samples.

Look at data and see count: table()
```{r}
x = table(wisc.df$diagnosis)
```
There are `r x[1]` benign and `r x[2]` malignant  
or There are `r x["B"]` benign an `r x["M"]` malignant

```{r}
wisc.data = as.matrix(wisc.df[,3:32])
row.names(wisc.data) <- wisc.df$id
```

```{r}
colnames(wisc.df)
```

```{r}
# vaule = TRUE, will show what is there, instead of just index
grep("_mean", colnames(wisc.df), value=TRUE)
```

Count how many features with _mean
```{r}
length(grep("_mean", colnames(wisc.df), value=TRUE))
```

```{r}
# use round to visualize easier! round second arg is how many numbers to see
round(apply(wisc.data,2,sd), 3)
```

Do PCA, scale is TRUE
```{r}
wisc.pr <- prcomp(wisc.data, scale=TRUE)
summary(wisc.pr)
```

without color
```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,2])
```
with color
```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,2], col= wisc.df$diagnosis, xlab="PC1", ylab="PC2")
```

```{r}
y = summary(wisc.pr)
y$importance[,"PC1"]
```
The first PC caputres `r y$importance[2,"PC1"] *100` % of variance

```{r}
data.scaled = scale(wisc.data)
data.dist <- dist(data.scaled)
wisc.hclust <- hclust(data.dist, method="complete")
plot(wisc.hclust)
abline(h=20, col="red", lty=2)
```

```{r}
wisc.hclust.clusters = cutree(wisc.hclust, k=4)
```
```{r}
table(wisc.hclust.clusters, wisc.df$diagnosis)
```

```{r}
dist1 = dist(wisc.pr$x[,1:7])
wisc.pr.hclust = hclust(dist1, method="ward.D2")
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
table(grps, wisc.df$diagnosis)
```
Can see false postie/negative, sensitivity, specificity....etc

```{r}
plot(wisc.pr$x[,1:2], col=grps)
```

```{r}
plot(wisc.pr$x[,1:2], col=wisc.df$diagnosis)
```

```{r}
g <- as.factor(grps)
levels(g)

```

```{r}
g <- relevel(g,2)
levels(g)
```
Re-plot with re-ordered grp
```{r}
plot(wisc.pr$x[,1:2], col=g)
```



```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```
```{r}
plot(wisc.pr$x[,1:2], col=g)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

