---
title: "Class14"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup Bioconductor DESeq2

# Do not run this code 
```{r eval=FALSE} 
install.packages("BiocManager")
BiocManager::install()

BiocManager::install("DESeq2")
```

## Data for today's class

```{r}
counts = read.csv("data/airway_scaledcounts.csv", stringsAsFactors = FALSE)
metadata = read.csv("data/airway_metadata.csv", stringsAsFactors = FALSE)
# genes are not factors (categorical)
```

How many genes do we have in the dataset?
```{r}
nrow(counts)
```

How many experiments?
```{r}
ncol(counts) - 1
```


# minus 1
```{r}
colnames(counts)[-1] 
```

```{r}
metadata$id
```

Make sure experiment names match in counts data and metadata 
```{r}
all(colnames(counts)[-1] == metadata$id)
```

Find the control id
```{r}
# metadata[metadata[,"dex"] == "control"]
# need to specify both row and col for subset
control.id = metadata[metadata$dex == "control", ]$id
```

Focus on the counts for control experiments
```{r}
head(counts[,control.id])
```

Mean counts for control exps
```{r}
control.mean = rowSums(counts[,control.id])/length(control.id)
names(control.mean) = counts$ensgene
```

# Do the same for treatment


```{r}
treated.id = metadata[metadata$dex == "treated", ]$id
treated.mean = rowSums(counts[,treated.id])/length(treated.id)
names(treated.mean) = counts$ensgene
```

Let's store them together
```{r}
meancounts = data.frame(control.mean, treated.mean)
```

Plot control vs treatment
```{r}
plot(meancounts$control.mean, meancounts$treated.mean)
```

Plot on log scale
```{r}
plot(meancounts$control.mean, meancounts$treated.mean, log="xy")
```
## Fold change in log
```{r}
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
head(meancounts)
```

Remove zero
```{r}
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)
```

Test removing zero with test dataframe
```{r}
x = data.frame(happy=c(4,2,0,0), sad=c(2,0,0,1))
```

```{r}
x == 0
```

```{r}
which(x==0, arr.ind=T)
```

```{r}
which(x==0, arr.ind=T)[,1]
```

```{r}
indx = unique(which(x==0, arr.ind=T)[,1])
#arr.ind > make into array of row and col, not just a vector going thru columns

# select row with zero
x[indx,]
```

Select row with no zeros
```{r}
x[-indx,]
```

## Continure where we left off and select data with no zero
```{r}
to.rm = which(meancounts[,1:2]==0, arr.ind=TRUE)[,1]
mycounts = meancounts[-to.rm,]
head(mycounts)
```

How many genes do I have left after removing rows with zero?
```{r}
nrow(mycounts)
```

```{r}
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)
```

Use sum to count TRUE and FALSE (0 and 1)
```{r}
sum(up.ind)
sum(down.ind)
```

## Add annotation, install and load packages
```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")
```


See what is available in org.Hs
```{r}
columns(org.Hs.eg.db)
```



Use mapIds to add annotation, specify key - column
```{r}
mycounts$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(mycounts), # Our genenames
                     keytype="ENSEMBL",        # The format of our genenames
                     column="SYMBOL",          # The new format we want to add
                     multiVals="first")
```

```{r}
head(mycounts)
```

# DESeq2 analysis
```{r}
library(DESeq2)
```


Setup our object for DESeq analysis
```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex, 
                              tidy=TRUE)
dds
```

```{r}
dds = DESeq(dds)
```

```{r}
res = results(dds)
res
```

# Volcano plot
```{r}
plot(res$log2FoldChange, res$padj)
```

```{r}
plot(res$log2FoldChange, -log(res$padj))
```

Color our plot
```{r}
# Setup our custom point color vector 
mycols <- rep("gray", nrow(res))
mycols[ abs(res$log2FoldChange) > 2 ]  <- "red" 

inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

# Volcano plot with custom colors 
plot( res$log2FoldChange,  -log(res$padj), 
 col=mycols, ylab="-Log(P-value)", xlab="Log2(FoldChange)" )

# Cut-off lines
abline(v=c(-2,2), col="gray", lty=2)
abline(h=-log(0.1), col="gray", lty=2)
```

